### 2021-02-09 docs/manuscript.tex
move methods to method section, add some points to discussion

    Similarly, there are also HLA-allele independent effects like antigen processing that influence presentation on MHC I. Second generational tools like MHCFlurry 2.0~\cite{ODonnell2020} use both binding affinity and antigen processing data to improve epitope prediction. For our tool HAMdetector, we use the output of MHCFlurry to benefit from this binding affinity and antigen processing data in order to predict HLA-associated mutations.

@@

  Recent advantages in Bayesian inference include so-called sparsity-promoting priors. Sparsity-promoting priors convey the a priori expectation that most coefficients in a regression model are close to 0. This assumption also applies to HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes. If no epitope spanning a given alignment position is presented on the MHC I molecule, any association between a replacement and the respective HLA allele  is likely to be due to random variation alone (or HLA linkage disequilibrium).

@@

@@

\begin{itemize}
  \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by X and were also used in the Phylogenetic Dependency Network study~\cite{Carlson2008}. \hl{DaHa: Wer ist für diese Sequenzen der Ansprechpartner?}
  \item A set of 351 HIV sequences mostly spanning the Pol gene from the Arevir database~\cite{Roomp2006}.
  \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm.
  \item A set of 104 Hepatitis-D-Virus sequences that were kindly provided by Michael Roggendorf.
  \item A set of 41 HIV sequences that were kindly provided by Rongge Yang.
\end{itemize}

@@

\hl{DaHo: $\hat{R}$ und $n_{\mathrm{eff}}$ sollten in den Methods eingeführt werden}
For all the Bayesian models, Markov Chain Monte Carlo sampling was done using Stan~\cite{Stan2021}. All model fits showed no signs of inference issues. In total, 4 different chains with 1000 samples each were drawn from the posterior. The effective sample size for all model parameters exceeded 500, R-hat convergence diagnostic values were below 1.01 in all cases.

\subsection*{Posterior predictive checks}
\hl{DaHo: Sollte in den Methods eingeführt werden}
Bayesian modelling allows a simple yet effective strategy to test models: By simulating data under the model's assumptions and then comparing simulated and observed data, model misspecifications can be identified and highlight possible model improvements.

Here, we compare the calibration of the posterior predictive probability of observing the replacement to the actual observed data using calibration plots.
The plots are similar to the one showed in figure~\ref{fig:phylogeny-calibration} but this time do not evaluate the probabilities based on the phylogenetic model alone, but show the calibration of the posterior predictive probabilities based on the full model.

An example calibration plot of the HAMdetector model on the Core protein sequence data of the HBV dataset is shown in figure~\ref{fig:calibration}. Plots for the other models and datasets are shown in the supplementary.

@@

All models show a strong enrichment of HLA-associated mutations compared to a randomly permutated list of HLA allele - replacement pairs (gray lines). Fisher's exact test performs about as well as the simple logistic regression model, as from a statistical perspective, they are both similar.
The horseshoe prior alone is a drastic improvement over Fisher's exact test and the logistic regression model with non-sparsifying priors, even though it does not include any external information. The benefit comes from including the assumption that most regression coefficients are expected to be very close to 0.
The logistic regression model with horseshoe prior works roughly as well as the Phylogenetic Dependency Network, which includes much more information.
Note that the line for the Phylogenetic Dependency Network is shorter because it only outputs associations with a q-value lower than 0.2 by default.

@@

\subsection{Leave-one-out cross-validation}
\hl{DaHo: LOO auch in Methods einführen}
In addition to testing the model against experimentally validated epitopes, we also performed leave-one-out cross-validation. While the comparison against known epitopes compares inferences of the regression coefficients, leave-one-out cross-validation compares different models in regard to their predictive performance, i.e. how well they can predict the presence or absence of a replacement in a sequence based on the available data.

As the compared models are all special cases of the complete model, we can use cross-validation to gain insight on the relative importance of each aspect of the model.

@@

Predictive accuracy increases as the models include more and more information.
Note that this is not due to overfitting, because leave-one-out cross-validation approximates the model performance on unseen data. Additional tables are shown in the supplementary, the results are consistent across all datasets.

Note that the model with horseshoe prior alone already has a much higher elpd than the standard logistic regression model, even though it does not use any additional external data. This is because including the sparsity assumption allows the model to better separate signal from noise and the uncertainty of the close-to-zero coefficients does not propagate into uncertainty of the estimated y.

@@

\subsection{Generalization}
no overfitting but improvements show for different data sets and different viruses
supplementary figures

@@

The results show low posterior probabilities for most of the false-positives and high posterior probabilities for most replacements that could be experimentally validated in IFN gamma production assays.
One important exception is P89T~-~HLA-B*37, which was experimentally validated but does not have strong support in our model.

@@

if quality is the same for all viruses: differences between immune/virus interactions become visible

recombinant forms

4-digit partial pooling

@@

\hl{DaHo: wir haben zwar mehrere Datensätze, gehen aber praktisch nicht auf die Resultate mit diesen Datensätzen ein. Wie besprochen sollte dieses Manuskript eher methodisch sein, d.h.~es ist ok, dass wir nicht *viel* auf die virologische/immunologische Seite eingehen. Aber, dass wir überhaupt nichts zu den Resultaten sagen, ist etwas zu wenig. Z.B. könnte man die Vielfalt der Datensätze nutzen um den Punkt zu machen, dass die Verbesserungen nicht von den Details des Virus abhängen. Vielleicht fällt Dir ja noch was dazu ein.}

>>>>

    Similarly, there are also HLA-allele independent effects like antigen processing that influence presentation on MHC I. Second generational tools like MHCFlurry 2.0~\cite{ODonnell2020} use both binding affinity and antigen processing data to improve epitope prediction. For our tool HAMdetector, we use the output of MHCFlurry to benefit from this data in order to predict HLA-associated mutations.

@@

  Recent advantages in Bayesian inference include so-called sparsity-promoting priors~\cite{Piironen2017}. Sparsity-promoting priors convey the a priori expectation that most coefficients in a regression model are close to 0. This assumption also applies to HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes. If no epitope spanning a given alignment position is presented on the MHC I molecule, any association between a replacement and the respective HLA allele  is likely to be due to random variation alone (or HLA linkage disequilibrium).

@@

  \subsection{Model diagnostics}

  \textit{Posterior predictive checks}

  Here, we use these so-called posterior predictive checks to compare the calibration of the predicted probabilities of observing the replacement to the actual observed data.

  \textit{Convergence diagnostics}

  We use Stan (version 2.23)~\cite{Stan2021} to specify and fit our Bayesian models.
  Stan provides an implementation of a dynamic Hamilitonian Monte Carlo sampler, which is a powerful technique to approximate expectations with respect to the posterior distribution~\cite{Betancourt2017}. One advantage of using such methods is that they provide sensitive diagnostics to identify a wide range of possible sampling issues.
  We use the split-\(\hat{R}\) convergence diagnostic~\cite{Gelman1992} to identify convergence issues. We require a value of \(\hat{R}\) below 1.01 for all model parameters. Additionally, we require that the effective sample size \(N_{\text{eff}}\) is above 500 for all model parameters and that sampling occurs without any divergent transitions. A detailed explanation of these quantities is provided in the Stan reference manual.

  \textit{Leave-one-out cross-validation}

  In addition to testing our model on real-world data, we also performed leave-one-out cross-validation to examine the relative importance of the additions of epitope prediction, sparsity and phylogeny with respect to predictive performance.
  For this, we create several variations of the model that are all special cases of the full HAMdetector model.
  By comparing the model performance of these variations, we can gain insight on the relative importance of each aspect of the model.

@@

\begin{itemize}
  \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by the Zabrina Brumme and were also used in the Phylogenetic Dependency Network study~\cite{Carlson2008}.
  \item A set of 351 HIV sequences mostly spanning the Pol gene from the Arevir database~\cite{Roomp2006}.
  \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm.
  \item A set of 104 Hepatitis-D-Virus sequences that were kindly provided by Michael Roggendorf.
  \item A set of 41 HIV sequences that were kindly provided by Rongge Yang.
\end{itemize}

@@

 All model fits showed no signs of inference issues. In total, samples were drawn from 4 different chains with 1000 iterations each. The effective sample size exceeded 500 for all model parameters, \(\hat{R}\) convergence diagnostic values were below 1.01 in all cases.

\subsection*{Posterior predictive checks}

Figure~\ref{fig:calibration} compares the calibration of the posterior predictive probability of observing the replacement to the actual observed data for the Core protein of the HBV dataset. Plots for the other models and datasets are shown in the supplementary.

The plots are similar to the one showed in figure~\ref{fig:phylogeny-calibration}, however, they do not evaluate the probabilities based on the phylogenetic model alone, but instead show the calibration of the posterior predictive probabilities based on the full model.

@@

All models show a strong enrichment of HLA-associated mutations compared to a randomly permuted list of HLA allele - replacement pairs (gray lines). Fisher's exact test performs about as well as the simple logistic regression model, as from a statistical perspective, they are both similar.
The horseshoe prior alone is a drastic improvement over Fisher's exact test and the logistic regression model with non-sparsifying priors, even though it does not include any external information. The benefit comes from including the assumption that most regression coefficients are expected to be very close to 0.
The logistic regression model with horseshoe prior works roughly as well as the Phylogenetic Dependency Network, which includes much more information.
Note that the line for the Phylogenetic Dependency Network is shorter because it only outputs associations with a q-value lower than 0.2 by default.

@@

\subsection{Leave-one-out cross-validation}

@@

Predictive accuracy increases as the models include more and more information.
Note that the model with horseshoe prior alone already has a much higher elpd than the standard logistic regression model, even though it does not use any additional external data. This is because including the sparsity assumption allows the model to better separate signal from noise and the uncertainty of the close-to-zero coefficients does not propagate into uncertainty of the estimated y.

@@

\subsection{Generalization}

The models were applied on sequence data from a variety of different viruses and show a considerable improvement over currently applied methods that are consistent across datasets.
These results are not due to overfitting, because leave-one-out cross-validation approximates the model performance on unseen data.

@@

The results show low posterior probabilities for most of the false-positives and high posterior probabilities for most replacements that could be experimentally validated in IFN-\(\gamma\) production assays.
One important exception is P89T~-~HLA-B*37, which was experimentally validated but does not have strong support in our model.

@@

@@

If the quality of the prediction is the same for all viruses, as it appears to be in our results, this method can be used to probe the frequency of HLA immune escape across different viruses.

In further research we plan to use HAMdetector to study circulating recombinant forms in HIV. One possible explanation for the occurrence of these recombinant forms is improved fitness by evading several immune epitopes at once.

While our model is an improvement over existing methods, a further addition that we plan to implement is partial pooling across 4-digit HLA alleles. It has recently been appreciated that binding specificities can vary drastically across  HLA alleles from the same allele group. For predicting HLA associated mutations, there has consequently been a shift to use 4-digit resolution data whenever available.

This is not without downsides however, because overall, binding specificities are more similar for alleles in the same group, and, therefore, treating them as completely separate might unnecessarily fragment the available data.

In Bayesian statistics, it is not required to chose one extreme or the other, With partial pooling, estimates from the same allele group do share some information across each other, but they are also allowed to vary if necessary.
Therefore, implementing partial pooling might be another possibility to further increase the reliability of computational methods to study HLA escape.
